---
title: "MISCADA ASML: Classification Summative Coursework Part 1"
output: html_notebook
---

The dataset used will be heart failure. 
       
Libraries
```{r}
library(tidyverse)
library(ggplot2)
library(reshape2)
library(skimr)
library(ggfortify)
library(GGally)
library(mlr3verse)
library(data.table)
library(dplyr)
library(kableExtra)
library(pROC)
```

Load the Data
```{r}
data <- readr::read_csv("heart_failure.csv")
```

Viewing the Data 

```{r}
View(data)
```

skim() - it tells the variable types, how many values are missing, and a mini histogram for each variable
```{r}
skim(data) 

```

Also check the basic structure and confirm there are no missing values:
```{r}
str(data)  #understand data structure
summary(data)  #summary stats
sum(is.na(data)) #any missing values?
sapply(data, is.factor) #is anyything a factor?

cat("Anaemia percentage:", mean(data$anaemia) * 100, "%\n")
cat("Diabetes percentage:", mean(data$diabetes) * 100, "%\n")
cat("High Blood Pressure percentage:", mean(data$high_blood_pressure) * 100, "%\n")
cat("Smoking percentage:", mean(data$smoking) * 100, "%\n")

```


Feature Engineering
```{r}
data <- data |> #binary to readable labels 
  mutate(
    anaemia_label     = case_when(anaemia == 1 ~ "Anaemic",            TRUE ~ "Not Anaemic"),
    diabetes_label    = case_when(diabetes == 1 ~ "Diabetic",          TRUE ~ "Not Diabetic"),
    high_bp_label     = case_when(high_blood_pressure == 1 ~ "High BP", TRUE ~ "Normal BP"),
    smoking_label     = case_when(smoking == 1 ~ "Smoker",             TRUE ~ "Non-Smoker"),
    outcome           = case_when(fatal_mi == 1 ~ "Fatal Event",       TRUE ~ "Survived"),
    comorbidity_count = anaemia + diabetes + high_blood_pressure + smoking,
    age_group         = case_when(
      age < 50 ~ "Under 50",
      age < 65 ~ "50-64",
      age < 75 ~ "65-74",
      TRUE     ~ "75+"
    )
  )

data |> select(anaemia, anaemia_label, fatal_mi, outcome, comorbidity_count, age_group) |> head(10) #check its correct
```

Examining Fatality Rates by Group
```{r}
fatality_by_age <- data |>
  group_by(age_group) |>
  summarise(total = n(), fatal_events = sum(fatal_mi), pct_fatal = fatal_events / total * 100)

ggplot(fatality_by_age |> arrange(desc(pct_fatal)),
       aes(x = reorder(age_group, -pct_fatal), y = pct_fatal, fill = age_group)) +
  geom_col() +
  geom_text(aes(label = paste0(round(pct_fatal, 1), "%")), vjust = -0.5, size = 4) +
  labs(title = "Fatality Rate by Age Group",
       subtitle = "Older patients show substantially higher mortality",
       x = "Age Group", y = "Fatal Event Rate (%)") +
  theme_minimal() + guides(fill = "none")
```

Fatality Rate by others
```{r}
fatality_by_comorbidity <- data |>
  group_by(comorbidity_count) |>
  summarise(total = n(), fatal_events = sum(fatal_mi), pct_fatal = fatal_events / total * 100)

ggplot(fatality_by_comorbidity,
       aes(x = factor(comorbidity_count), y = pct_fatal, fill = factor(comorbidity_count))) +
  geom_col() +
  geom_text(aes(label = paste0(round(pct_fatal, 1), "%\n(n=", total, ")")), vjust = -0.3, size = 3.5) +
  labs(title = "Fatality Rate by Number of Comorbidities",
       subtitle = "Anaemia + Diabetes + High BP + Smoking combined",
       x = "Number of Comorbidities", y = "Fatal Event Rate (%)") +
  theme_minimal() + guides(fill = "none") +
  expand_limits(y = max(fatality_by_comorbidity$pct_fatal) * 1.15)
```

Further Graphical Interpretations
```{r}
ggplot(data, aes(x = age, fill = outcome)) +
  geom_density(alpha = 0.5) +
  labs(title = "Age Distribution by Survival Status", x = "Age", y = "Density", fill = "Outcome") +
  theme_minimal()

ggplot(data, aes(x = outcome, fill = outcome)) +
  geom_bar() +
  geom_text(stat = 'count', aes(label = after_stat(count)), vjust = -0.5, size = 5) +
  labs(title = "Total Count of Outcomes", x = "Patient Outcome", y = "Number of Patients") +
  theme_minimal() + expand_limits(y = 220) + guides(fill = "none")

ggplot(data, aes(x = smoking_label, fill = outcome)) +
  geom_bar(position = "dodge") +
  labs(title = "Survival Count by Smoking Status", x = "Smoking Status", y = "Count", fill = "Outcome") +
  scale_fill_manual(values = c("Survived" = "skyblue", "Fatal Event" = "tomato")) +
  theme_minimal()

ggplot(data, aes(x = serum_creatinine, y = ejection_fraction, color = outcome)) +
  geom_point(alpha = 0.7, size = 3) +
  scale_color_manual(values = c("Survived" = "skyblue", "Fatal Event" = "tomato")) +
  labs(title = "Survival by Serum Creatinine and Ejection Fraction",
       subtitle = "High risk: low ejection fraction and high creatinine (bottom-right)",
       x = "Serum Creatinine (mg/dL)", y = "Ejection Fraction (%)") +
  theme_minimal() +
  geom_vline(xintercept = 1.3, linetype = "dashed", color = "grey") +
  geom_hline(yintercept = 30,  linetype = "dashed", color = "grey")


```

Pairwise Relationships
```{r}
ggpairs(data |> select(age, ejection_fraction, serum_creatinine, serum_sodium, outcome),
        aes(color = outcome, alpha = 0.5)) +
  scale_color_manual(values = c("Survived" = "skyblue", "Fatal Event" = "tomato")) +
  scale_fill_manual(values = c("Survived" = "skyblue", "Fatal Event" = "tomato")) +
  labs(title = "Pairwise Relationships of Key Clinical Variables") +
  theme_minimal()
```

Correlation Matrix
```{r}
corr_mat <- round(cor(data |> select(where(is.numeric))), 2)
melted_corr_mat <- reshape2::melt(corr_mat)
ggplot(data = melted_corr_mat, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "steelblue", mid = "white", high = "tomato", midpoint = 0) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "All Numeric Variables", fill = "Correlation")
```

Preparing Data for Modelling
```{r}
data_model <- data |>
  select(-anaemia_label, -diabetes_label, -high_bp_label, -smoking_label,
         -outcome, -age_group, -time)

skim(data_model)
names(data_model)#final set of predictors to use
```

Principal Component Analysis 
```{r}
pca_input <- data_model[, -which(names(data_model) == "fatal_mi")]#all cols but fatal_mi
pca_result <- prcomp(pca_input, center = TRUE, scale = TRUE)

summary(pca_result)
data$fatal_mi_factor <- factor(data$fatal_mi, labels = c("Survived", "Fatal Event"))#dummy variable to explain fatal_mi as a factor for PCA
plot(pca_result, type = "l", main = "PCA Variance")#View the loadings (weights) for the first 4 components
round(pca_result$rotation[, 1:4], 3)
```

Advanced PCA analysis 
```{r}
data_model_pca <- data_model |>
  mutate(log_creatinine = log1p(serum_creatinine),
         log_cpk        = log1p(creatinine_phosphokinase)) |>
  select(-serum_creatinine, -creatinine_phosphokinase)

selected_vars     <- c("age", "ejection_fraction", "serum_sodium", "log_creatinine", "log_cpk")
pca_input_reduced <- data_model_pca[, selected_vars]
pca_reduced       <- prcomp(pca_input_reduced, center = TRUE, scale = TRUE)

autoplot(pca_reduced,
         data   = data_model_pca |> mutate(fatal_mi_factor = factor(fatal_mi, labels = c("Survived", "Fatal Event"))),
         colour = 'fatal_mi_factor', loadings = TRUE, loadings.label = TRUE) +
  scale_color_manual(values = c("Survived" = "skyblue", "Fatal Event" = "tomato")) +
  labs(title = "Reduced & Log-Transformed PCA",
       subtitle = "Focusing on top clinical predictors", colour = "Outcome") +
  theme_minimal()

summary(pca_reduced)
plot(pca_reduced, type = "l", main = "Reduced PCA Variance")
round(pca_reduced$rotation[, 1:4], 3)
```
```{r}
var_full <- pca_result$sdev^2
var_reduced <- pca_reduced$sdev^2
plot(var_full, type = "b", pch = 16, lwd = 2,
     xlab = "Principal Component",
     ylab = "Variance",
     main = "PCA Variance Comparison")

lines(var_reduced, type = "b", pch = 17, col = "red", lwd = 2)
legend("topright",
       legend = c("Full PCA", "Reduced PCA"),
       col = c("black", "red"),
       lwd = 2,
       pch = c(16, 17))
```

