---
title: "MISCADA ASML: Classification Summative Coursework"
output: html_notebook
---



Libraries
```{r}
library(tidyverse)
library(ggplot2)
library(reshape2)
library(skimr)
library(ggfortify)
library(GGally)
library(mlr3verse)
library(data.table)
library(dplyr)
library(kableExtra)
library(knitr)
library(pROC)
library(glmnet)
```

Load the Data
```{r}
data <- readr::read_csv("heart_failure.csv")
```

Viewing the Data 
```{r}
#View(data)
```

skim() = it tells the variable types, how many values are missing, and a mini histogram for each variable
```{r}
skim(data) 
# All 13 variables are numeric doubles no factors, no missing values.

```


Also check the basic structure and confirm there are no missing values:
```{r}
str(data)#understand data structure
summary(data) #summary stats
sum(is.na(data)) #any missing values?
sapply(data, is.factor) #is anyything a factor?

cat("Anaemia percentage:", mean(data$anaemia) * 100, "%\n")
cat("Diabetes percentage:", mean(data$diabetes) * 100, "%\n")
cat("High Blood Pressure percentage:", mean(data$high_blood_pressure) * 100, "%\n")
cat("Smoking percentage:", mean(data$smoking) * 100, "%\n")

```

Feature Engineering
```{r}
data <- data |>#binary to readable labels 
  mutate(
    anaemia_label     = case_when(anaemia == 1 ~ "Anaemic",            TRUE ~ "Not Anaemic"),
    diabetes_label    = case_when(diabetes == 1 ~ "Diabetic",          TRUE ~ "Not Diabetic"),
    high_bp_label     = case_when(high_blood_pressure == 1 ~ "High BP", TRUE ~ "Normal BP"),
    smoking_label     = case_when(smoking == 1 ~ "Smoker",             TRUE ~ "Non-Smoker"),
    outcome           = case_when(fatal_mi == 1 ~ "Fatal Event",       TRUE ~ "Survived"),
    existingconditions_count = anaemia + diabetes + high_blood_pressure + smoking,
    age_group         = case_when(
      age < 50 ~ "Under 50",
      age < 65 ~ "50-64",
      age < 75 ~ "65-74",
      TRUE     ~ "75+"
    )
  )

data |> select(anaemia, anaemia_label, fatal_mi, outcome, existingconditions_count, age_group) |> head(10) #check its correct
```


Examining Fatality Rates by Group
```{r}
fatality_by_age <- data |>
  group_by(age_group) |>
  summarise(total = n(), fatal_events = sum(fatal_mi), pct_fatal = fatal_events / total * 100)

ggplot(fatality_by_age |> arrange(desc(pct_fatal)),
       aes(x = reorder(age_group, -pct_fatal), y = pct_fatal, fill = age_group)) +
  geom_col() +
  geom_text(aes(label = paste0(round(pct_fatal, 1), "%")), vjust = -0.5, size = 4) +
  labs(title = "Fatality Rate by Age Group",
       subtitle = "Older patients show substantially higher mortality",
       x = "Age Group", y = "Fatal Event Rate (%)") +
  theme_minimal() + guides(fill = "none")
```

Fatality Rate by others
```{r}
fatality_by_existingconditions <- data |>
  group_by(existingconditions_count) |>
  summarise(total = n(), fatal_events = sum(fatal_mi), pct_fatal = fatal_events / total * 100)

ggplot(fatality_by_existingconditions,
       aes(x = factor(existingconditions_count), y = pct_fatal, fill = factor(existingconditions_count))) +
  geom_col() +
  geom_text(aes(label = paste0(round(pct_fatal, 1), "%\n(n=", total, ")")), vjust = -0.3, size = 3.5) +
  labs(title = "Fatality Rate by Number of Coexisting conditions",
       subtitle = "Anaemia + Diabetes + High BP + Smoking combined",
       x = "Number of Coexisting conditions", y = "Fatal Event Rate (%)") +
  theme_minimal() + guides(fill = "none") +
  expand_limits(y = max(fatality_by_existingconditions$pct_fatal) * 1.15)
```

Further Graphical Interpretations
```{r}
ggplot(data, aes(x = age, fill = outcome)) +
  geom_density(alpha = 0.5) +
  labs(title = "Age Distribution by Survival Status", x = "Age", y = "Density", fill = "Outcome") +
  theme_minimal()
```


Pairwise Relationships
```{r}
ggpairs(data |> select(age, ejection_fraction, serum_creatinine, serum_sodium, outcome),
        aes(color = outcome, alpha = 0.5)) +
  scale_color_manual(values = c("Survived" = "skyblue", "Fatal Event" = "tomato")) +
  scale_fill_manual(values = c("Survived" = "skyblue", "Fatal Event" = "tomato")) +
  labs(title = "Pairwise Relationships of Key Clinical Variables") +
  theme_minimal()
```

Correlation Matrix
```{r}
corr_mat <- round(cor(data |> select(where(is.numeric))), 2)
melted_corr_mat <- reshape2::melt(corr_mat)
ggplot(data = melted_corr_mat, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "steelblue", mid = "white", high = "tomato", midpoint = 0) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "All Numeric Variables", fill = "Correlation")
```


Principal Component Analysis 
```{r}
data_model <- data |>
  select(-anaemia_label, -diabetes_label, -high_bp_label, -smoking_label,
         -outcome, -age_group, -time)

skim(data_model)
names(data_model)#final set of predictors to use in pCA

pca_input <- data_model[, -which(names(data_model) == "fatal_mi")]#all cols but fatal_mi
pca_result <- prcomp(pca_input, center = TRUE, scale = TRUE)

summary(pca_result)#importance of each component
data$fatal_mi_factor <- factor(data$fatal_mi, labels = c("Survived", "Fatal Event"))#dummy variable to explain fatal_mi as a factor for PCA
#variance explained by each principal component
plot(pca_result, type = "l", main = "PCA Variance")#View the loadings (weights) for the first 4 components
round(pca_result$rotation[, 1:4], 3)
```

Advanced PCA analysis 
so log transform serum_creatinine and creatinine_phosphokinase for PCA different view
```{r}
#Log transform skewed variables
data_model_pca <- data_model |>
  mutate(log_creatinine = log1p(serum_creatinine),
         log_cpk        = log1p(creatinine_phosphokinase)) |>
  select(-serum_creatinine, -creatinine_phosphokinase)

selected_vars<- c("age", "ejection_fraction", "serum_sodium", "log_creatinine", "log_cpk")
pca_input_reduced <- data_model_pca[, selected_vars]
pca_reduced<- prcomp(pca_input_reduced, center = TRUE, scale = TRUE)

summary(pca_reduced)
plot(pca_reduced, type = "l", main = "Reduced PCA Variance")
round(pca_reduced$rotation[, 1:4], 3)


var_full<- pca_result$sdev^2
var_reduced<- pca_reduced$sdev^2
plot(var_full, type = "b", pch = 16, lwd = 2,
     xlab = "Principal Component",
     ylab = "Variance",
     main = "PCA Variance Comparison")

lines(var_reduced, type = "b", pch = 17, col = "red", lwd = 2)
legend("topright",
       legend = c("Full PCA", "Reduced PCA"),
       col = c("black", "red"),
       lwd = 2,
       pch = c(16, 17))
```

Data Preparation
```{r}
set.seed(100)
data <- readr::read_csv("heart_failure.csv")

#RF is scale invariant so uses data_model as usual
data_model <- data |>
  mutate(
    existingconditions_count = anaemia + diabetes + high_blood_pressure + smoking,
    fatal_mi          = factor(fatal_mi, levels = c(1, 0),
                               labels = c("Fatal", "Survived"))
  ) |>
  select(-time)

#log transform right skewed continuous variables
data_model_lr <- data_model |>
  mutate(
    creatinine_phosphokinase = log1p(creatinine_phosphokinase),#13x ratio
    serum_creatinine         = log1p(serum_creatinine),#7x ratio
    platelets                = log1p(platelets) #3x ratio
  )

skim(data_model)
```

Train / Test Split
```{r}
n <- nrow(data_model)
train_idx <- sample(n, size = floor(0.8 * n))

train_data    <- data_model[train_idx, ]
test_data     <- data_model[-train_idx, ]
train_data_lr <- data_model_lr[train_idx, ]
test_data_lr  <- data_model_lr[-train_idx, ]

cat("Training set size:", nrow(train_data), "\n")
cat("Test set size:    ", nrow(test_data),  "\n")
cat("Fatal event rate (train):", round(mean(train_data$fatal_mi == "Fatal") * 100, 1), "%\n")
cat("Fatal event rate (test): ", round(mean(test_data$fatal_mi  == "Fatal") * 100, 1), "%\n")
```

Task Definition
```{r}
#Fatal is positive class
#raw features for Random Forest
task_rf_hr <- TaskClassif$new(id= "HeartFailure",backend= train_data,target= "fatal_mi",positive= "Fatal")
#log-transformed for Logistic Regression
task_lr_hf <- TaskClassif$new(id = "HeartFailure_LR",backend= train_data_lr,target= "fatal_mi",positive = "Fatal")
```

Cross-Validation Strategy
```{r}
#5-fold CV once per task.
cv5 <- rsmp("cv", folds = 5)
cv5_lr <- rsmp("cv", folds = 5)
cv5$instantiate(task_rf_hr)
cv5_lr$instantiate(task_lr_hf)
```

Performance Measures
```{r}
measures <- list(
  msr("classif.ce"),# classification error
  msr("classif.acc"),# accuracy
  msr("classif.auc"),# AUC, 1 is perfect, worse then 0.5 bad 
  msr("classif.sensitivity"),#fatal events correctly found, identifies this asymmetry as a reason to move beyond raw accuracy.
  msr("classif.fnr"),#fatal events incorrectly missed
  msr("classif.fpr")#survivors incorrectly flagged
)
```



Model 1: Logistic Regression
```{r}
lrn_log_reg <- lrn("classif.log_reg", predict_type = "prob")
pl_log_reg  <- po("encode") %>>% po(lrn_log_reg)
```

Model 2: Random Forest
```{r}
lrn_ranger <- lrn("classif.ranger", predict_type = "prob", num.trees = 500,
                   importance = "impurity", keep.inbag = TRUE) # keep.inbag needed to extract each tree OOB curve
```

Baseline Classifier
```{r}
lrn_baseline <- lrn("classif.featureless", predict_type = "prob")
```


Benchmark for the models
```{r}
res_baseline<- resample(task_rf_hr,    lrn_baseline, cv5,    store_models = TRUE) #all on 5 fold
res_lr<- resample(task_lr_hf, pl_log_reg,   cv5_lr, store_models = TRUE)
res_rf<- resample(task_rf_hr,    lrn_ranger,   cv5,    store_models = TRUE)

baseline_res<- as.data.frame(t(res_baseline$aggregate(measures))) |> mutate(Model = "Baseline")
lr_res<- as.data.frame(t(res_lr$aggregate(measures)))       |> mutate(Model = "Logistic Regression")
rf_res<- as.data.frame(t(res_rf$aggregate(measures)))       |> mutate(Model = "Random Forest")

benchkmark_results <- bind_rows(baseline_res, lr_res, rf_res) |> select(Model, everything())

kable(benchkmark_results, digits = 3, caption = "5-Fold CV Results") |>
  kable_styling(full_width = FALSE, position = "center")
```


OOB Error Curve sto show  Random Forest Early Stopping Diagnostic
```{r}
rf_oob <- ranger::ranger( #once it flattens, additional trees add no meaningful variance reduction.
  formula    = fatal_mi ~ ., data= train_data, num.trees = 500, mtry= floor(sqrt(ncol(train_data) - 1)), keep.inbag = TRUE,
  probability = TRUE,importance = "impurity"
)

#extract by refitting progressively using a sequence of tree counts
oob_trees  <- c(1, seq(10, 500, by = 10))
oob_errors <- sapply(oob_trees, function(b) {
  m <- ranger::ranger(formula = fatal_mi ~ ., data = train_data,
                      num.trees = b, probability = TRUE, verbose = FALSE)
  m$prediction.error
})

oob_rf_df <- data.frame(trees = oob_trees, oob_error = oob_errors)
ggplot(oob_rf_df, aes(x = trees, y = oob_error)) +
  geom_line(colour = "steelblue", linewidth = 0.8) +
  geom_hline(yintercept = min(oob_rf_df$oob_error),
             linetype = "dashed", colour = "tomato", linewidth = 0.8) +
  labs(title = "Random Forest OOB Error Curve — Fold 1 (500 Trees)",
       subtitle = "Dashed = minimum OOB error. Plateau around 100",
       x = "Number of Trees", y = "OOB Classification Error") +
  theme_minimal()
```

Random Forest Variable Importance
```{r}
#for each feature, sum the Gini impurity reduction
#High importance means the feature was frequently useful
rf_imp <- res_rf$learners[[1]]$model$variable.importance
importance_df <- data.frame(
  variable   = names(rf_imp),
  importance = as.numeric(rf_imp)
) |> arrange(desc(importance))

print(importance_df)
ggplot(importance_df, aes(x = reorder(variable, importance), y = importance)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Random Forest Variable Importance — Fold 1",
       subtitle = "Gini impurity reduction summed across all 500 trees",
       x = NULL, y = "Mean Decrease in Gini Impurity") +
  theme_minimal()
```

Logistic Regression Coefficients
```{r}
#fitted glm from fold 1, positive coefficient means increases log-odds of a Fatal event
lr_fold1_model <- res_lr$learners[[1]]$graph_model$pipeops$classif.log_reg$state$model

coef_df <- data.frame(
  variable   = names(coef(lr_fold1_model)),
  coef       = as.numeric(coef(lr_fold1_model)),
  odds_ratio = as.numeric(exp(coef(lr_fold1_model)))
) |>
  filter(variable != "(Intercept)") |>
  arrange(desc(abs(coef)))

print(round(coef_df[, c("coef", "odds_ratio")], 4))

ggplot(coef_df, aes(x = reorder(variable, coef), y = coef, fill = coef > 0)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "tomato", "FALSE" = "steelblue"),
                    labels = c("TRUE" = "Increases Fatal risk", "FALSE" = "Decreases Fatal risk")) +
  labs(title = "Logistic Regression Coefficients — Fold 1",
       subtitle = "Red = increases log-odds of Fatal; Blue = decreases",
       x = NULL, y = "Coefficient", fill = NULL) +
  theme_minimal()
```

default models evaluated on test set
```{r}
# Retrain both models on the full training set for test set evaluation
default_lrn_lr <- po("encode") %>>% po(lrn("classif.log_reg", predict_type = "prob"))
default_lrn_lr$train(task_lr_hf)

default_lrn_rf <- lrn("classif.ranger", predict_type = "prob", num.trees = 500, importance = "impurity")
default_lrn_rf$train(task_rf_hr)

task_test_rf_obj <- TaskClassif$new(id = "HeartFailure_test",    backend = test_data,    target = "fatal_mi", positive = "Fatal")
task_test_lr_obj <- TaskClassif$new(id = "HeartFailure_test_LR", backend = test_data_lr, target = "fatal_mi", positive = "Fatal")

predicted_lr_test <- default_lrn_lr$predict(task_test_lr_obj)[[1]]
predicted_rf_test <- default_lrn_rf$predict(task_test_rf_obj)
```

Default model comparison
```{r}
cat("Default model test comparison \n")
metrics <- c("classif.auc", "classif.sensitivity", "classif.ce", "classif.fnr", "classif.fpr")

for (m in metrics) {
  lr_val <- predicted_lr_test$score(msr(m))
  rf_val <- predicted_rf_test$score(msr(m))
  winner <- if ((m %in% c("classif.auc", "classif.sensitivity", "classif.acc")) &&
                rf_val > lr_val) "RF" else
             if ((m %in% c("classif.ce", "classif.fnr", "classif.fpr")) &&
                 rf_val < lr_val) "RF" else "LR"
  cat(sprintf("  %-30s  LR: %.3f   RF: %.3f   [%s]\n", m, lr_val, rf_val, winner))
}

cat("\nLogistic Regression:\n")
predicted_lr_test$confusion

cat("\nRandom Forest:\n")
predicted_rf_test$confusion
cat("\nWinner: Random Forest since sensitivity means missing a fatal error\n")

#Random Forest has the higher AUC so better overall discrimination. Logistic Regression has the higher sensitivity and catches more.
#In a medical context where missing a fatal event is the worst error, sensitivity is the clinically critical metric like in Lecture 9
```



Logistic Regression Regularisation
```{r}
#alpha = 1 Lasso does automatic feature selection, 
#alpha = 0 Ridge all predictors
#alpha = 0.5 Elastic Net does both
using_logreg_plain  <- lrn("classif.log_reg", predict_type = "prob", id = "logreg_plain")
using_glmnet_l1     <- lrn("classif.glmnet",  predict_type = "prob", alpha = 1,   lambda = 0.01,  id = "lasso")
using_glmnet_l2     <- lrn("classif.glmnet",  predict_type = "prob", alpha = 0,   lambda = 0.01,  id = "ridge")
using_glmnet_enet   <- lrn("classif.glmnet",  predict_type = "prob", alpha = 0.5, lambda = 0.01,  id = "elasticnet")
using_glmnet_strong <- lrn("classif.glmnet",  predict_type = "prob", alpha = 0.5, lambda = 0.05,  id = "glmnet_strong")
using_glmnet_weak   <- lrn("classif.glmnet",  predict_type = "prob", alpha = 0.5, lambda = 0.001, id = "glmnet_weak")

res_logreg <- benchmark(data.table(
  task       = list(task_lr_hf),
  learner    = list(
    po("encode") %>>% po(using_logreg_plain),
    po("encode") %>>% po(using_glmnet_l1),
    po("encode") %>>% po(using_glmnet_l2),
    po("encode") %>>% po(using_glmnet_enet),
    po("encode") %>>% po(using_glmnet_strong),
    po("encode") %>>% po(using_glmnet_weak)
  ),
  resampling = list(cv5_lr)
), store_models = TRUE)

logReg_tuning <- res_logreg$aggregate(measures) |>
  as.data.frame() |>
  select(learner_id, classif.auc, classif.sensitivity, classif.ce, classif.fnr) |>
  arrange(desc(classif.auc))

kable(logReg_tuning, digits = 3,
      caption = "Logistic Regression - Regularisation Variants (5-fold CV)") |>
  kable_styling(full_width = FALSE, position = "center")
```

Variables removed abive 
```{r}
# Variables shrunk to zero contributed no independent signal after accounting for other predictors
findMissingVars <- tryCatch(
  res_logreg$resample_result(2)$learners[[1]]$model$classif.glmnet$model,
  error = function(e) NULL
)

if (is.null(findMissingVars)) {
  X_lasso <- model.matrix(fatal_mi ~ . - 1, data = train_data_lr)
  y_lasso <- as.numeric(train_data_lr$fatal_mi == "Fatal")
  findMissingVars <- glmnet(X_lasso, y_lasso, family = "binomial", alpha = 1, lambda = 0.01)
  cat("Lasso coefficients (standalone refit) -- zero = excluded:\n")
} else {
  cat("Lasso coefficients (Fold 1) -- zero = excluded:\n")
}
print(round(as.matrix(coef(findMissingVars)), 4))
```

Best logistic regression variant
```{r}
best_lr_id <- logReg_tuning$learner_id[1]
cat("Best LR variant=", best_lr_id, "\n")
cat("AUC         =", round(logReg_tuning$classif.auc[1], 3), "\n")
cat("Sensitivity =", round(logReg_tuning$classif.sensitivity[1], 3), "\n")
cat("Error       =", round(logReg_tuning$classif.ce[1], 3), "\n")
```





Random Forest Hyperparameter Tuning
given it won above, more time spent on this tuning then the Logistic reg

by using number of trees
```{r}
#More trees reduce variance by averaging more bootstrap samples

using_100_trees <- lrn("classif.ranger", predict_type = "prob", num.trees = 100,  id = "rf_100")
using_300_trees  <- lrn("classif.ranger", predict_type = "prob", num.trees = 300,  id = "rf_300")
using_500_trees  <- lrn("classif.ranger", predict_type = "prob", num.trees = 600,  id = "rf_500")
using_1000_trees <- lrn("classif.ranger", predict_type = "prob", num.trees = 1000, id = "rf_1000")

#adding trees to a Random Forest never causes overfitting
testing_trees <- benchmark(data.table(
  task       = list(task_rf_hr),
  learner    = list(using_100_trees, using_300_trees, using_500_trees, using_1000_trees),
  resampling = list(cv5)
), store_models = TRUE)

trees_results <- testing_trees$aggregate(measures) |>
  as.data.frame() |>
  select(learner_id, classif.auc, classif.sensitivity, classif.ce, classif.fnr) |>
  arrange(desc(classif.auc))

kable(trees_results, digits = 3,
      caption = "num.trees") |>
  kable_styling(full_width = FALSE, position = "center")

best_trees     <- trees_results$learner_id[1]
best_trees_val <- as.integer(gsub("rf_", "", best_trees))
cat("\nBest num.trees:", best_trees_val, "| AUC:", round(trees_results$classif.auc[1], 3), "\n")
```



mtry = how many features are considered at each split 
```{r}
using_mtry2 <- lrn("classif.ranger", predict_type = "prob", num.trees = best_trees_val, mtry = 2, id = "rf_mtry2")
using_mtry3 <- lrn("classif.ranger", predict_type = "prob", num.trees = best_trees_val, mtry = 3, id = "rf_mtry3")
using_mtry5 <- lrn("classif.ranger", predict_type = "prob", num.trees = best_trees_val, mtry = 5, id = "rf_mtry5")
using_mtry8 <- lrn("classif.ranger", predict_type = "prob", num.trees = best_trees_val, mtry = 8, id = "rf_mtry8")

testing_mtry <- benchmark(data.table(
  task       = list(task_rf_hr),
  learner    = list(using_mtry2, using_mtry3, using_mtry5, using_mtry8),
  resampling = list(cv5)
), store_models = TRUE)

mtry_results <- testing_mtry$aggregate(measures) |>
  as.data.frame() |>
  select(learner_id, classif.auc, classif.sensitivity, classif.ce, classif.fnr) |>
  arrange(desc(classif.auc))

kable(mtry_results, digits = 3,
      caption = "mtry") |>
  kable_styling(full_width = FALSE, position = "center")

best_mtry_row <- mtry_results$learner_id[1]
best_mtry_val <- as.integer(gsub("rf_mtry", "", best_mtry_row))
cat("\nBest mtry:", best_mtry_val, "| AUC:", round(mtry_results$classif.auc[1], 3), "\n")
```



min.node.size to set the minimum number of observations required in a leaf node
```{r}
using_node1  <- lrn("classif.ranger", predict_type = "prob", num.trees = best_trees_val, mtry = best_mtry_val, min.node.size = 1,  id = "rf_node1")
lrn_rf_node5  <- lrn("classif.ranger", predict_type = "prob", num.trees = best_trees_val, mtry = best_mtry_val, min.node.size = 5,  id = "rf_node5")
using_node10 <- lrn("classif.ranger", predict_type = "prob", num.trees = best_trees_val, mtry = best_mtry_val, min.node.size = 10, id = "rf_node10")
using_node15 <- lrn("classif.ranger", predict_type = "prob", num.trees = best_trees_val, mtry = best_mtry_val, min.node.size = 15, id = "rf_node15")
using_node20 <- lrn("classif.ranger", predict_type = "prob", num.trees = best_trees_val, mtry = best_mtry_val, min.node.size = 20, id = "rf_node20")

tesing_node <- benchmark(data.table(
  task       = list(task_rf_hr),
  learner    = list(using_node1, lrn_rf_node5, using_node10, using_node15, using_node20),
  resampling = list(cv5)
), store_models = TRUE)

node_results <- tesing_node$aggregate(measures) |>
  as.data.frame() |>
  select(learner_id, classif.auc, classif.sensitivity, classif.ce, classif.fnr) |>
  arrange(desc(classif.auc))

kable(node_results, digits = 3,
      caption = "min.node.size") |>
  kable_styling(full_width = FALSE, position = "center")

best_node_row <- node_results$learner_id[1]
best_node_val <- as.integer(gsub("rf_node", "", best_node_row))
cat("\nBest min.node.size:", best_node_val, "| AUC:", round(node_results$classif.auc[1], 3), "\n")
```

max.depth = applies a hard cap on tree depth from the top down
```{r}
using_d3  <- lrn("classif.ranger", predict_type = "prob", num.trees = best_trees_val, mtry = best_mtry_val, min.node.size = best_node_val, max.depth = 3,  id = "rf_depth3")
using_d5  <- lrn("classif.ranger", predict_type = "prob", num.trees = best_trees_val, mtry = best_mtry_val, min.node.size = best_node_val, max.depth = 5,  id = "rf_depth5")
using_d10 <- lrn("classif.ranger", predict_type = "prob", num.trees = best_trees_val, mtry = best_mtry_val, min.node.size = best_node_val, max.depth = 10, id = "rf_depth10")
using_dul <- lrn("classif.ranger", predict_type = "prob", num.trees = best_trees_val, mtry = best_mtry_val, min.node.size = best_node_val, id = "rf_unlimited")

testing_depth <- benchmark(data.table(
  task       = list(task_rf_hr),
  learner    = list(using_d3, using_d5, using_d10, using_dul),
  resampling = list(cv5)
), store_models = TRUE)

depth_results <- testing_depth$aggregate(measures) |>
  as.data.frame() |>
  select(learner_id, classif.auc, classif.sensitivity, classif.ce, classif.fnr) |>
  arrange(desc(classif.auc))

kable(depth_results, digits = 3,
      caption = "max.depth") |>
  kable_styling(full_width = FALSE, position = "center")

best_depth_row <- depth_results$learner_id[1]
best_depth_val <- if (best_depth_row == "rf_unlimited") NULL else
  as.integer(gsub("rf_depth", "", best_depth_row))
cat("\nBest max.depth:", ifelse(is.null(best_depth_val), "unlimited", best_depth_val),
    "| AUC:", round(depth_results$classif.auc[1], 3), "\n")
```



Final model configuration vs default models
```{r}
set.seed(100)
training_rf_default <- lrn("classif.ranger", predict_type = "prob", num.trees = 500,
                        importance = "impurity", id = "rf_default")

#best from the above
rf_hyperparams_tuned <- lrn("classif.ranger", predict_type = "prob",
                     num.trees = best_trees_val, mtry = best_mtry_val,
                     min.node.size = best_node_val, importance = "impurity", id = "rf_tuned")

#add max.depth if a finite value found above
if (!is.null(best_depth_val)) rf_hyperparams_tuned$param_set$values$max.depth <- best_depth_val

bothmodels <- benchmark(data.table(
  task       = list(task_rf_hr),
  learner    = list(training_rf_default, rf_hyperparams_tuned),
  resampling = list(cv5)
), store_models = TRUE)

results_def_tuned <- bothmodels$aggregate(measures) |>
  as.data.frame() |>
  select(learner_id, classif.auc, classif.sensitivity, classif.ce, classif.fnr, classif.fpr)

kable(results_def_tuned, digits = 3, caption = "Tuned RF vs Default (5-fold CV)") |>
  kable_styling(full_width = FALSE, position = "center")

cat("\n")
cat(sprintf("Default AUC = %.3f\n", results_def_tuned$classif.auc[results_def_tuned$learner_id == "rf_default"]))
cat(sprintf("Tuned RF AUC= %.3f\n", results_def_tuned$classif.auc[results_def_tuned$learner_id == "rf_tuned"]))
cat(sprintf("AUC improvement= %.3f\n", diff(results_def_tuned$classif.auc[c(1,2)])))
```


Retrain final tuned models on full training set
predicted_rf and predicted_lr are the final test set predictions used for performance metrics
```{r}
final_random_forest_model <- lrn("classif.ranger",
                 predict_type  = "prob",
                 num.trees= best_trees_val,
                 mtry = best_mtry_val,
                 min.node.size= best_node_val,
                 importance= "impurity")
if (!is.null(best_depth_val)) final_random_forest_model$param_set$values$max.depth <- best_depth_val
final_random_forest_model$train(task_rf_hr)

final_log_regression_model <- 
  po("encode") %>>% 
  po("learner",
     lrn("classif.glmnet",
         predict_type = "prob",
         alpha = 0.5,
         lambda = 0.05,
         id = "glmnet_strong"))

final_log_regression_model$train(task_lr_hf)

task_test_rf <- TaskClassif$new(id = "test_rf", backend = test_data,    target = "fatal_mi", positive = "Fatal")
task_test_lr <- TaskClassif$new(id = "test_lr", backend = test_data_lr, target = "fatal_mi", positive = "Fatal")

predicted_rf <- final_random_forest_model$predict(task_test_rf)
predicted_lr <- final_log_regression_model$predict(task_test_lr)[[1]]
```

Picking Threshold
```{r}
thresholds <- seq(0.05, 0.95, by = 0.025) #go across many thresholds

sweep_rf <- map_dfr(thresholds, function(t) {
  predicted_rf$set_threshold(t)
  data.frame(
    threshold   = t,
    sensitivity = predicted_rf$score(msr("classif.sensitivity")),
    specificity = predicted_rf$score(msr("classif.specificity")),
    fpr         = predicted_rf$score(msr("classif.fpr")),
    fnr         = predicted_rf$score(msr("classif.fnr")),
    error       = predicted_rf$score(msr("classif.ce")),
    ppv         = tryCatch(predicted_rf$score(msr("classif.ppv")), error = function(e) NA)
  )
})
predicted_rf$set_threshold(0.5)

sweep_lr <- map_dfr(thresholds, function(t) {
  predicted_lr$set_threshold(t)
  data.frame(
    threshold   = t,
    sensitivity = predicted_lr$score(msr("classif.sensitivity")),
    specificity = predicted_lr$score(msr("classif.specificity")),
    fpr         = predicted_lr$score(msr("classif.fpr")),
    fnr         = predicted_lr$score(msr("classif.fnr")),
    error       = predicted_lr$score(msr("classif.ce")),
    ppv         = tryCatch(predicted_lr$score(msr("classif.ppv")), error = function(e) NA)
  )
})
predicted_lr$set_threshold(0.5)

#where sensitivity >= 0.6 AND fpr <= 0.25, minimise FNR
target_sensitivity <- 0.60
target_fpr<- 0.25

best_threshold_rf <- sweep_rf |>
  filter(sensitivity >= target_sensitivity, fpr <= target_fpr) |>
  arrange(fnr) |>
  slice(1) |>
  pull(threshold)
#if no threshold meets both targets
if (length(best_threshold_rf) == 0) {
  best_threshold_rf <- sweep_rf |>
    filter(sensitivity >= 0.50) |>
    arrange(fnr) |>
    slice(1) |>
    pull(threshold)
}

best_threshold_lr <- sweep_lr |>
  filter(sensitivity >= target_sensitivity, fpr <= target_fpr) |>
  arrange(fnr) |>
  slice(1) |>
  pull(threshold)
if (length(best_threshold_lr) == 0) {
  best_threshold_lr <- sweep_lr |>
    filter(sensitivity >= 0.50) |>
    arrange(fnr) |>
    slice(1) |>
    pull(threshold)
}

cat("Best RF threshold :", best_threshold_rf, "\n")
cat("Best LR threshold :", best_threshold_lr, "\n")

```

```{r}
ggplot(sweep_rf, aes(x = threshold)) +
  geom_line(aes(y = sensitivity, colour = "Sensitivity (TPR)"),  linewidth = 1.1) +
  geom_line(aes(y = fpr,         colour = "False Positive Rate"), linewidth = 1.1) +
  geom_line(aes(y = fnr,         colour = "False Negative Rate"), linewidth = 1.1, linetype = "dashed") +
  geom_vline(xintercept = 0.5,             linetype = "dotted",   colour = "grey40",    linewidth = 0.7) +
  geom_vline(xintercept = best_threshold_rf, linetype = "longdash", colour = "darkorange", linewidth = 0.8) +
  annotate("text", x = 0.52,                   y = 0.95,
           label = "Default\n(0.5)", size = 3, colour = "grey40") +
  annotate("text", x = best_threshold_rf + 0.02, y = 0.95,
           label = paste0("Chosen\n(", best_threshold_rf, ")"), size = 3, colour = "darkorange") +
  scale_colour_manual(values = c("Sensitivity (TPR)"  = "steelblue",
                                 "False Positive Rate" = "tomato",
                                 "False Negative Rate" = "darkgreen")) +
  labs(title = "Threshold Analysis for Random Forest",
       subtitle = "Lowering threshold increases sensitivity and FPR; raises specificity and FNR",
       x = "Classification Threshold", y = "Rate", colour = NULL) +
  theme_minimal() + theme(legend.position = "bottom")
```

Performance at threshold
```{r}
predicted_rf$set_threshold(0.5)
default_metrics <- data.frame(
  Metric  = c("Sensitivity", "FPR", "FNR", "Specificity", "Error"),
  Default = c(round(predicted_rf$score(msr("classif.sensitivity")), 3),
              round(predicted_rf$score(msr("classif.fpr")),         3),
              round(predicted_rf$score(msr("classif.fnr")),         3),
              round(predicted_rf$score(msr("classif.specificity")), 3),
              round(predicted_rf$score(msr("classif.ce")),          3))
)

predicted_rf$set_threshold(best_threshold_rf)
chosen_metrics <- data.frame(
  Chosen = c(round(predicted_rf$score(msr("classif.sensitivity")), 3),
             round(predicted_rf$score(msr("classif.fpr")),         3),
             round(predicted_rf$score(msr("classif.fnr")),         3),
             round(predicted_rf$score(msr("classif.specificity")), 3),
             round(predicted_rf$score(msr("classif.ce")),          3))
)

threshold_comparison <- bind_cols(default_metrics, chosen_metrics)

kable(threshold_comparison, digits = 3,
      caption = paste0("Random Forest: Default (0.5) vs Chosen (", best_threshold_rf, ")")) |>
  kable_styling(full_width = FALSE, position = "center") |>
  row_spec(c(1, 3), bold = TRUE, background = "#fff3cd")

cat(sprintf("\nRF confusion matrix at chosen threshold = %s :\n", best_threshold_rf))
predicted_rf$confusion
```


ROC Curves
```{r}
# ROC curves show the full discrimination profile across all thresholds.
# The closer the curve hugs the top-left corner, the better the model.
# The diagonal represents random guessing (AUC = 0.5).
predicted_rf$set_threshold(best_threshold_rf)
predicted_lr$set_threshold(best_threshold_lr)

roc_rf <- roc(response  = as.numeric(predicted_rf$truth == "Fatal"),
              predictor = predicted_rf$prob[, "Fatal"], quiet = TRUE)
roc_lr <- roc(response  = as.numeric(predicted_lr$truth == "Fatal"),
              predictor = predicted_lr$prob[, "Fatal"], quiet = TRUE)

roc_df <- bind_rows(
  data.frame(fpr = 1 - roc_rf$specificities, tpr = roc_rf$sensitivities,
             model = sprintf("Random Forest (AUC = %.3f)", auc(roc_rf))),
  data.frame(fpr = 1 - roc_lr$specificities, tpr = roc_lr$sensitivities,
             model = sprintf("Logistic Regression (AUC = %.3f)", auc(roc_lr)))
)

ggplot(roc_df, aes(x = fpr, y = tpr, colour = model)) +
  geom_line(linewidth = 1.2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey60", linewidth = 0.6) +
  annotate("text", x = 0.6, y = 0.1, label = "Random classifier (AUC = 0.5)",
           colour = "grey50", size = 3.2) +
  scale_colour_manual(values = c("tomato", "steelblue")) +
  labs(title = "ROC Curves with Final Models on Test Set",
       subtitle = "Top-left corner = perfect discrimination. Diagonal = random.",
       x = "False Positive Rate (1 - Specificity)", y = "True Positive Rate (Sensitivity)",
       colour = NULL) +
  theme_minimal() + theme(legend.position = "bottom")
```




Final Test Set 
```{r}
# Compare tuned RF against default RF on the test set before selecting threshold
final_default_rf <- lrn("classif.ranger", predict_type = "prob", num.trees = 500, importance = "impurity")
final_default_rf$train(task_rf_hr)
pred_default <- final_default_rf$predict(task_test_rf)

test_metrics <- c("classif.auc", "classif.sensitivity", "classif.ce", "classif.fnr", "classif.fpr")

test_comparison <- data.frame(
  Metric  = test_metrics,
  Default = sapply(test_metrics, function(m) pred_default$score(msr(m))),
  Tuned   = sapply(test_metrics, function(m) {
    predicted_rf$set_threshold(0.35)
    predicted_rf$score(msr(m))
  })
)

kable(test_comparison, digits = 3,
      caption = "Final Test Set: Default vs Tuned RF") |>
  kable_styling(full_width = FALSE, position = "center")

cat("\nDefault RF confusion matrix:\n")
pred_default$confusion

cat("\nTuned RF confusion matrix:\n")
predicted_rf$set_threshold(best_threshold_rf)
predicted_rf$confusion
```


Full Performance Metrics — final models at chosen thresholds
```{r}
predicted_rf$set_threshold(best_threshold_rf)
predicted_lr$set_threshold(best_threshold_lr)

all_metrics <- c(
  "classif.auc",
  "classif.sensitivity",
  "classif.specificity",
  "classif.ce",
  "classif.acc",
  "classif.fnr",
  "classif.fpr",
  "classif.ppv",  # precision of those flagged fatal how many are fatal
  "classif.npv"   # negative predictive value of those flagged survived are they survived
)

metric_names <- c(
  "AUC","Sensitivity (Recall)","Specificity","Classification Error","Accuracy","False Negative Rate","False Positive Rate","Precision (PPV)","Negative Predictive Value (NPV)"
)

perf_table <- data.frame(
  RF = sapply(all_metrics, function(m)
    tryCatch(round(predicted_rf$score(msr(m)), 3), error = function(e) NA)),
  LR = sapply(all_metrics, function(m)
    tryCatch(round(predicted_lr$score(msr(m)), 3), error = function(e) NA)),
  Perfect = c(1, 1, 1, 0, 1, 0, 0, 1, 1)
)

rownames(perf_table) <- metric_names

kable(perf_table, digits = 3,
      caption = "Full Performance Metrics in Test Set") |>
  kable_styling(full_width = FALSE, position = "center") |>
  row_spec(which(metric_names %in% c("AUC", "Sensitivity (Recall)")),
           bold = TRUE, background = "#f0f8ff") |>
  column_spec(2, bold = TRUE)
```


Precision-Recall Analysis
```{r}
PrecisionRecall_df <- map_dfr(thresholds, function(t) {
  predicted_rf$set_threshold(t)
  data.frame(threshold = t,
             precision = tryCatch(predicted_rf$score(msr("classif.ppv")),         error = function(e) NA),
             recall    = tryCatch(predicted_rf$score(msr("classif.sensitivity")), error = function(e) NA),
             model     = "Random Forest")
})

#Logistic regression PR for comparison
PrecisionRecall_df_lr <- map_dfr(thresholds, function(t) {
  predicted_lr$set_threshold(t)
  data.frame(threshold = t,
             precision = tryCatch(predicted_lr$score(msr("classif.ppv")),         error = function(e) NA),
             recall    = tryCatch(predicted_lr$score(msr("classif.sensitivity")), error = function(e) NA),
             model     = "Logistic Regression")
})

PrecisionRecalls <- bind_rows(PrecisionRecall_df, PrecisionRecall_df_lr) |> filter(!is.na(precision), !is.na(recall))


ggplot(PrecisionRecalls, aes(x = recall, y = precision, colour = model)) +
  geom_line(linewidth = 1.1) +
  geom_hline(yintercept = 0.32, linetype = "dashed", colour = "grey50", linewidth = 0.6) +
  annotate("text", x = 0.1, y = 0.34, label = "No-skill baseline (prevalence = 0.32)",
           colour = "grey50", size = 3) +
  scale_colour_manual(values = c("tomato", "steelblue")) +
  labs(title = "Precision-Recall Curves with Final Models on Test Set",
       subtitle = "Above the dashed line = better than predicting Fatal for every patient",
       x = "Recall (Sensitivity)", y = "Precision (Positive Predictive Value)", colour = NULL) +
  xlim(0, 1) + ylim(0, 1) +
  theme_minimal() + theme(legend.position = "bottom")
```


Final Configuration Summary 
```{r}
predicted_rf$set_threshold(best_threshold_rf)

cat("Final Metrics \n")
cat("AUC        :", round(predicted_rf$score(msr("classif.auc")),         3), "\n")
cat("Sensitivity:", round(predicted_rf$score(msr("classif.sensitivity")), 3), "\n")
cat("Error      :", round(predicted_rf$score(msr("classif.ce")),          3), "\n")
cat("FNR        :", round(predicted_rf$score(msr("classif.fnr")),         3), "\n")
cat("FPR        :", round(predicted_rf$score(msr("classif.fpr")),         3), "\n")

cat("\nFinal Random Forest matrix:\n")
predicted_rf$confusion

cat("\nFinal Random Forest Model \n")
cat(sprintf("  num.trees     : %d\n",   best_trees_val))
cat(sprintf("  mtry          : %d\n",   best_mtry_val))
cat(sprintf("  min.node.size : %d\n",   best_node_val))
cat(sprintf("  max.depth     : %s\n",
            ifelse(is.null(best_depth_val), "unlimited", best_depth_val)))
cat("  importance    : impurity (Gini)\n")
cat("  Threshold     :", best_threshold_rf, "\n")
```

